---
output:
  pdf_document: default
html_document: default
editor_options: 
  chunk_output_type: console
---
  
```{r include=FALSE}
knitr::opts_chunk$set(echo = T)
set.seed(123)

is_interactive <- FALSE
```


```{r echo=FALSE, results='hide', message=FALSE, echo = F}

library(dplyr)
library(magrittr)
library(tidyr)
library("scatterplot3d")
library(magrittr)
library(languageR)
library(ggplot2)
theme_set(theme_bw())

library(rgl)
library(knitr)
knit_hooks$set(webgl = hook_webgl)

# 
# labelled_arrow <- function(p, x_start, y_start, x_end, y_end, x_label, y_label, label, color, arrowhead_size = unit(0.2, "cm")) {
#     p + geom_segment(data=NULL, aes(x=x_start, y=y_start, xend=x_end, yend=y_end), 
#                         color = color, arrow = arrow(length = arrowhead_size), size = 1) + 
#           geom_text(x = x_label, y = y_label, label = label, color = color)
# }
# 
# labelled_slope_arrow <- function(p, x_start, y_start, x_end, y_end, 
#                                  x_label, y_label, label, color, 
#                                  arrowhead_size = unit(0.2, "cm"))
# {
#     p <- p + geom_segment(data=NULL, aes(x=x_start, y=y_start, xend=x_end, yend=y_start), color = color, size = 1)
#     p %>% labelled_arrow(x_start = x_end, y_start = y_start, x_end = x_end, y_end = y_end, 
#                          x_label = x_label, y_label = y_label, label = label, color = color)
# }

source("./code/create_cab_data.R")

```



# Examples

## The Dative Verbs Data

- I want to know if the length of the theme and the recipient influence the proportion of recipient-theme word orders, and by how much. (In this sample!)

- Let's prepare the data.
```{r}

# load all packages that will be required
library(dplyr)
library(magrittr)
library(ggplot2)
library(languageR)

# create a data frame with the required columns 
dative_relevant <- dative %>% dplyr::select(RealizationOfRecipient, LengthOfRecipient, LengthOfTheme)

# take a look
head(dative_relevant, 3)

# discretize length of recipient 
dative_relevant %<>% mutate( cat_LengthOfRecipient = ifelse(LengthOfRecipient > median(LengthOfRecipient) , "long", "short") %>% as.factor() )
dative_relevant %<>% mutate( cat_LengthOfTheme = ifelse(LengthOfTheme > median(LengthOfTheme) , "long", "short") %>% as.factor() )

```



### By Length of Recipient

- Now, let's summarize and plot it
```{r}
# compute percentage of default word orders by categorical recipient length
(dative_summary1 <- dative_relevant %>% 
  group_by(cat_LengthOfRecipient) %>% 
  summarize(perc_def = mean(RealizationOfRecipient == "NP"), 
            N = n()))
# plot averages
ggplot(dative_summary1, aes(x= cat_LengthOfRecipient, perc_def)) + geom_point()
```

- Now, let's fit a mode with treatment contrasts and look at its coefficients.
```{r}
# Fit model 1A (treatment contrasts):
# Fit a linear model with treatment contrasts according to the following specification:
#  perc_def = a + b*length of recipient 

# create a treatment contrast for (categorical) length ('ct' is for 'contrast, treatment')
dative_summary1 %<>% mutate(ct_LR = dplyr::recode(cat_LengthOfRecipient, "short"=0, "long"=1))

# use coef(lm(...)) to fit the linear model described above and extract its coefficients
coef(lm( perc_def ~ 1 + ct_LR, data = dative_summary1))
```

- Now, let's fit a mode with sum contrasts and look at its coefficients.
```{r}
# Fit model 1B (sum contrasts):
# Fit a linear model with sum contrasts according to the following specification:
#  perc_def = a + b*length of recipient 

# create a treatment contrast for (categorical) length ('cs' is for 'contrast, sum')
dative_summary1 %<>% mutate(cs_LR = dplyr::recode(cat_LengthOfRecipient, "short"=-.5, "long"=.5))

# use coef(lm(...)) to fit the linear model described above and extract its coefficients
coef(lm( perc_def ~ 1 + cs_LR, data = dative_summary1))
```

### By Length of Theme

- Now, let's summarize and plot the averages.
```{r}
# compute percentage of default word orders by categorical theme length
(dative_summary2 <- dative_relevant %>% 
                      group_by(cat_LengthOfTheme) %>% 
                      summarize(perc_def = mean(RealizationOfRecipient == "NP"), 
                                N = n()))
# plot averages
ggplot(dative_summary2, aes(x= cat_LengthOfTheme, perc_def)) + geom_point()
```

- Now, let's fit a mode with treatment contrasts and look at its coefficients.
```{r}
# Fit model 2A (treatment contrasts):
# Fit a linear model with treatment contrasts according to the following specification:
#  perc_def = a + b*length of theme 

# create a treatment contrast for (categorical) length ('ct' is for 'contrast, treatment')
dative_summary2 %<>% mutate(ct_LT = dplyr::recode(cat_LengthOfTheme, "short"=0, "long"=1))

# use coef(lm(...)) to fit the linear model described above and extract its coefficients
coef(lm( perc_def ~ 1 + ct_LT, data = dative_summary2))
```

- Now, let's fit a mode with sum contrasts and look at its coefficients.
```{r}
# Fit model 1B (sum contrasts):
# Fit a linear model with sum contrasts according to the following specification:
#  perc_def = a + b*length of recipient 

# create a treatment contrast for (categorical) length ('ct' is for 'contrast, treatment')
dative_summary2 %<>% mutate(cs_LR = dplyr::recode(cat_LengthOfTheme, "short"=-.5, "long"=.5))

# use coef(lm(...)) to fit the linear model described above and extract its coefficients
coef(lm( perc_def ~ 1 + cs_LR, data = dative_summary1))
```


### By Length of Recipient and Length of Theme

- Now, let's summarize and plot the averages.

```{r}
# compute percentage of default word orders by categorical theme and recipient length
(dative_summary3 <-
  dative_relevant %>% group_by(cat_LengthOfTheme, cat_LengthOfRecipient) %>% 
  summarize(perc_def = mean(RealizationOfRecipient == "NP"), 
            N = n() ))

# plot averages
dative_summary3 %>% ggplot(aes( cat_LengthOfRecipient, perc_def, color = cat_LengthOfTheme, group = cat_LengthOfTheme)) + geom_point() + geom_line()
```

- Now, let's fit a mode with treatment contrasts and look at its coefficients.
```{r}
# Fit model 2A (treatment contrasts):
# Fit a linear model with treatment contrasts according to the following specification:
#  perc_def = a + b1*length of theme + b2*length of recipient + b3*length of theme*length of recipient

# create a treatment contrast for (categorical) length ('ct' is for 'contrast, treatment')
dative_summary3 %<>% mutate( ct_LR = dplyr::recode(cat_LengthOfRecipient, "short"=0, "long"=1),
                             ct_LT = dplyr::recode(cat_LengthOfTheme,     "short"=0, "long"=1)
                            )

# use coef(lm(...)) to fit the linear model described above and extract its coefficients
coef(lm( perc_def ~ 1 + ct_LT + ct_LR + ct_LT:ct_LR, data = dative_summary3))
```

- Please verify the meaning of the coefficients in the above plots, data frames and other models
  * **Intercept:** The `perc_def` value for the baseline value (i.e., short theme and short recipient). 
  * **Main effect**, coefficient for `ct_LT`: The effect of theme length when the recipient is short.
  * **Main effect**, coefficient for `ct_LR`: The effect of recipient length when the theme is short.
  * **Interaction**, coefficient for `ct_LT:ct_LR`: The *additional* effect of theme length when the the recipient is long. (*or:* The *additional* effect of **recipient** length when the the **theme** is long.)  


- Now, let's fit a mode with sum contrasts and look at its coefficients.
```{r}
# Fit model 1B (sum contrasts):
# Fit a linear model with sum contrasts according to the following specification:
#  perc_def = a + b1*length of theme + b2*length of recipient + b3*length of theme*length of recipient

# create a sum contrast for (categorical) length ('cs' is for 'contrast, sum')
dative_summary3 %<>% mutate( cs_LR = dplyr::recode(cat_LengthOfRecipient, "short"=-.5, "long"=.5),
                             cs_LT = dplyr::recode(cat_LengthOfTheme, "short"=-.5, "long"=.5)
)

# use coef(lm(...)) to fit the linear model described above and extract its coefficients
coef(lm( perc_def ~ 1 + ct_LT + ct_LR + ct_LT:ct_LR, data = dative_summary3))
```

- Please verify the meaning of the coefficients in the above plots, data frames and other models:
  * **Intercept:** Average percentage of default word orders.
  * **Main effect**, coefficient for `ct_LT`: Average effect of length of the theme.
  * **Main effect**, coefficient for `ct_LR`: Average effect of length of the recipient.
  * **Interaction**, coefficient for `ct_LT:ct_LR`: Change in the effect of `ct_LR` due to one unit of change in `ct_LT`. (*or:* Change in the effect of `ct_LT` due to one unit of change in `ct_LR`.)  



<!--
## The Sleep Data
  
- I want to know if bigger animals sleep more.
  
```{r eval=F}

mammalian_sleep <- 
  read.csv("./data/msleep_ggplot2.csv") %>% 
  dplyr::select(name, sleep_total, bodywt) %>%
  dplyr::rename(sleep_total_h = sleep_total, bodywt_kg = bodywt) %>%
  dplyr::mutate(sleep_total_h = round(sleep_total_h) )

head(mammalian_sleep)

ggplot(mammalian_sleep, aes(bodywt_kg, sleep_total_h)) + geom_point()
```
-->